---
title: "DATA607 - Week 12 Homework"
author: "Mike Dehn"
date: "12/08/2021"
output:
  pdf_document:
    latex_engine : pdflatex
    number_sections: true
---

```{r, setup, include=FALSE}
library(gamair)
data(chicago)
X <- chicago
head(X)
X1 <- na.omit(X)
X1 <- scale(X1)
dim(X1)
head(X1)
chi.omit <- 200
set.seed(1234)
in.row <- sample(719, chi.omit)
in.col <- sample(1:7, chi.omit, replace = TRUE)
X1.omit <- X1
index.omit <- cbind(in.row, in.col)
X1.omit[index.omit] <- NA
# Your missing data matrix is then X1.omit
```

# Iterative Matrix Completion Implementation

```{r}
Mat.Complete <- function(X, thresh, maxiter) {
  # Calculate the mean of each column so it can be used for initialization of
  # the imputed matrix
  x_bar <- colMeans(X, na.rm = TRUE)

  # Copy the original matrix into x_hat and replace all NA values with the mean
  # of its column
  x_hat <- X
  for (i in 1:ncol(X)) {
    x_hat[is.na(x_hat[, i]), i] <- x_bar[i]
  }

  # Helper variable that stores which matrix elements are missing
  ismiss <- is.na(X)

  # Initial state of the mean squared error used by the algorithm for tracking
  # improvement per iteration
  mssold <- mean((scale(X, x_bar, FALSE)[!ismiss])^2)
  mss0 <- mean(X[!ismiss]^2)

  # Key values that will track the loop conditions
  rel_err <- 1
  iter <- 0

  # Loop until our relative error is below the threshold value, or we reach the
  # maximum number of permitted iterations
  while (rel_err > thresh && iter < maxiter) {
    iter <- iter + 1

    # Algorithm 12.1, 2a - Perform principal component analysis (PCA) on the
    # approximation matrix. The resulting structure will provide matrix A and
    # matrix B in fields $x and $rotation, respectively.
    # NOTE: We use rank. = 1 for parity with the example in Section 12.5.2
    x_pca <- prcomp(x_hat, center = FALSE, rank. = 1)

    # Algorithm 12.1, 2b - Update approximation matrix to store a_i1 * b_i1 for
    # each of the missing elements in the original matrix
    x_app <- x_pca$x %*% t(x_pca$rotation)
    x_hat[ismiss] <- x_app[ismiss]

    # Algorithm 12.1, 2c - Calculate mean squared error for this iteration and
    # update the relative error
    mss <- mean(((X - x_app)[!ismiss])^2)
    rel_err <- (mssold - mss) / mss0
    mssold <- mss

    # Report on current state
    cat("Iter:", iter, "MSS:", mss, "Rel. Err:", rel_err, "\n")
  }

  # If we stopped due to number of iterations, warn the user
  if (rel_err > thresh) {
    warning("solution did not converge within permitted iterations")
  }

  # Return a list containing the completed matrix, the number of iterations,
  # and the relative error of the final iteration
  list(x_hat, iter, rel_err)
}
```

# Matrix Completion with `Mat.Complete`

```{r}
# Invoke matrix completion function on incomplete data matrix
thresh <- 1e-7
maxiter <- 30
x_imputed <- Mat.Complete(X1.omit, thresh, maxiter)
x_imputed[[1]]
```

# Matrix Completion with `softImpute`

```{r}
library(softImpute)

# Invoke softImpute matrix completion on incomplete data matrix. Use settings
# that best approximate what we did with the homebrew implementation
x_softImpute <- softImpute(X1.omit, thresh = thresh, maxit = maxiter,
                           trace.it = TRUE, type = "svd", rank.max = 1)
```

Based on the results above, we conclude that our original implementation of
iterative matrix completion has roughly equivalent output to that of the
`softImpute` package. Both algorithms complete with approximately the same
number of iterations and have similar relative errors upon their final
iteration.